\part[Nouvelles attaques sur \shaone]
    {New attacks on \shaone} 
\label{part:sha-1}

\section*{Overview}

The \shaone hash function was one of the first widely-deployed hash functions. It was designed by the NSA in 1995 as a modified version of 1993's \shazero, which itself
was more loosely based on 1990's \mdfour.

In 2005, Wang \etal presented the first major attack on \shaone, giving a method to find collisions for the hash function
in time equivalent to $\approx 2^{69}$ evaluations of the compression function, which is significantly faster than the $\approx 2^{80}$ required by a generic process.
However, the high cost of the attack implied that no explicit collision was computed at the time. This remains the case today, even if improvements to the attack
and ever more efficient hardware would make such an endeavour more affordable than it was in 2005.
The considerable amount of work that followed Wang \etal's original contribution then mostly focused on first getting a better understanding of the attack process, and
secondly on bringing innovations allowing to explicitly provide collisions for reduced versions of \shaone with the highest number of steps. 

Considering the efficient collision attacks developped since 2005, it may seem remarkable that \shaone
in fact still offers a comfortable security margin against first preimage attacks. Until our own work,
the attack reaching the highest number of steps was due to Knellwolf and Khovratovich, who showed in 2012 how to attack $57/80$ of the function, with an estimated
complexity of $2^{158.8}$ evaluations of the compression function.


\bigskip

The third and last part of this thesis describes new attacks on \shaone.

We start by presenting explicit collision attacks for the full compression function of \shaone. While this comes short of a practical
attack on the \emph{hash function}, this is the first explicit result on \shaone for a standard security notion. Furthermore, the efficient implementation framework that was developed
for this work could also be reused for hash function attacks, which have a very similar (and in some respect simpler) structure than our ``freestart'' case.

Collision attacks on \shaone \emph{à la} Wang have a rather specific structure, full of small technicalities. Thus, after briefly
introducting \shaone in \autoref{cha:shaone_pres}, we start with a presentation of this topic in \autoref{cha:shaone_hist},
that we hope to be self-contained. Our contributions on freestart collisions are presented next in \autoref{cha:shaone_new}.

We then conclude this thesis by presenting in \autoref{cha:shaone_pre}
the current best preimage attacks on \shaone, in terms of number of attacked steps. By extending the framework developped by Knellwolf and Khovratovich to
higher-order differential attacks, we are able to increase by five the number of steps of \shaone for which a preimage can be computed faster than with
a generic method.



\cleardoublepage
\chapter*{Contents}
\parttoc

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\chapter[Présentation de \shaone]
        {The \shaone hash function}
\label{cha:shaone_pres}

\input{../III_SHA-1/notations.tex}

\input{../III_SHA-1/sha1_presentation.tex}


\chapter[Une brève histoire des attaques en collision sur \shaone]
        {A brief history of collision attacks on \shaone}
\label{cha:shaone_hist}

\input{../III_SHA-1/history.tex}

\chapter[Collisions à initialisation libres pour \shaone]
        {Freestart collision attacks for \shaone}
\label{cha:shaone_new}

\input{../III_SHA-1/freestart.tex}

\input{../III_SHA-1/gpu.tex}

\input{../III_SHA-1/res76.tex}

\input{../III_SHA-1/res80.tex}

\FloatBarrier

\input{../III_SHA-1/conclusion.tex}

\chapter[Attaques en préimages sur \shaone]
        {Preimage attacks for \shaone}
\label{cha:shaone_pre}

\input{../III_SHA-1/preimages/Introduction.tex}

\input{../III_SHA-1/preimages/KKFramework.tex}

\input{../III_SHA-1/preimages/NewFramework.tex}

\input{../III_SHA-1/preimages/AppliSHA1.tex}


\FloatBarrier
