\part[Nouvelles attaques et constructions pour chiffres par bloc]
	{New attacks and elements of design for block ciphers}
\label{part:constructions}

\pagestyle{plain}
\section*{Overview}

A widespread design type for block ciphers is the \emph{SPN} structure. Initially, this stood for \emph{substitution-permutation network},
with \emph{permutation} referring to bit permutations; in its current usage, this term generally denotes arbitrary \ftwo-linear
mappings.

The rationale behind the SPN structure is that one may obtain a good block cipher round function by alternating a substitution
mapping made of the parallel application of \emph{S-boxes} (short for \emph{substitution boxes}) and a linear mapping. The role
of both mappings are complementary, and each is also usually selected to interact well with the other. While the S-boxes
are expected to break any exploitable linear and algebraic structure at the local level, the permutation
should spread the effects of the S-boxes at the global level, in particular to ensure that no local change in the state of the cipher remains so through
the entire cipher execution.

An advantage of SPNs over some other design structures is that they lend themselves relatively well to analysis, in particular with respect to
statistical attacks; this is exemplified by the \emph{wide-trail strategy} used in the design of the \aes. Another notable point
about SPNs is that they offer many possible tradeoffs in their instantiations. One can choose between large and small S-boxes,
large S-boxes being comparatively stronger than small ones but also more expensive; best-performing S-boxes or cheaper
average ones; bit permutations, particularly efficient in hardware, or non-permutation matrices; optimally diffusing (MDS)
matrices or not; state-wide matrices or not, etc. The design space of SPNs is thus quite large, and concrete instantiations
of the framework may be quite distinct one from another. 

\bigskip

In the second part of this thesis, we study two mappings intended to be used as part of an SPN round function, the second
of them being also instantiated in a complete block cipher design. On the cryptanalysis side, we start by presenting a simple attack of the
\proestotr authenticated-encryption scheme, using a related-key model.

In \autoref{cha:emrka}, we show how the simple related-key distinguisher on Even-Mansour constructions can be converted into a
key-recovery attack by considering related-key introduced with a modular addition rather than an XOR. Although this observation
is quite elementary, it finds a practical application in a very efficient related-key attack on \proestotr, which is an instantiation
of the OTR authenticated-encryption mode of operation with the \proest permutation.


In \autoref{cha:difmat}, we present large linear mappings with very good diffusion and efficient software implementations,
with respect to their sizes. They are derived from linear codes over a small field, typically $\Fst$, with a high dimension,
typically 16, and a high minimum distance. This results in
diffusion matrices with equally high dimension and a large branch number.
Because no MDS code is known to exist for the selected parameters, the matrices were derived from more flexible \emph{algebraic geometry} codes.
We present two simple yet efficient algorithms for the software implementation of matrix-vector multiplication in this context, and derive
conditions on the generator matrices of the codes to yield efficient encoders. We then specify an appropriate code and use its automorphisms as well as random sampling
to find good such matrices.

In \autoref{cha:littlun}, we present
the construction and implementation of an 8-bit S-box with a differential and linear branch number of 3.
We show an application by designing \littlunpride, a simple block cipher
based on bitsliced evaluations of the S-box and bit rotations which is designed to be efficient on 8-bit microcontrollers in particular.
The round function of \fly achieves good performance on its target platform, in terms of number of instructions per round and overall
code size, while providing good cryptographic properties.
The S-box also has an efficient implementation with SIMD instructions, a low implementation cost in hardware
and it can be masked efficiently thanks to its sparing use of non-linear gates and to the fact that it has a natural expression in
terms of a single 4-bit S-box.

\cleardoublepage
\chapter*{Contents}
\parttoc

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter[Attaques en clefs liées améliorées sur les schémas Even-Mansour]{Improved related-key attacks on Even-Mansour}
\pagestyle{ruled}
\label{cha:emrka}

\input{../I_PROLEGO/emrka.tex}


\chapter[Matrices de diffusion issues de codes géométriques]{Efficient diffusion matrices from algebraic geometry codes}
\label{cha:difmat}

\input{../II_LOTF/matrices/introduction}

\input{../II_LOTF/matrices/notations}

\input{../II_LOTF/matrices/agcodes}

\input{../II_LOTF/matrices/implem}

\input{../II_LOTF/matrices/diffag}

\input{../II_LOTF/matrices/applications}

\input{../II_LOTF/matrices/conclusion}

\input{../II_LOTF/matrices/appendix}

\FloatBarrier

\chapter[La boîte-S \littlun et le chiffre \fly]{The \littlun S-box and the \fly block cipher}
\label{cha:littlun}

\input{../II_LOTF/littlun/introduction}

\input{../II_LOTF/littlun/preliminaries}

\input{../II_LOTF/littlun/construction}

\input{../II_LOTF/littlun/implementation}

\input{../II_LOTF/littlun/application}

\FloatBarrier

\setcounter{section}{0}
\renewcommand\thesection{\Alph{section}}

\input{../II_LOTF/littlun/stats}
\FloatBarrier

\input{../II_LOTF/littlun/implemplus}

\input{../II_LOTF/littlun/avrimplem}
\FloatBarrier

\input{../II_LOTF/littlun/hwimplem}

\input{../II_LOTF/littlun/testvectors}

\input{../II_LOTF/littlun/maskedimplem}

\renewcommand\thesection{\arabic{section}}
