\section{Algebraic-geometry codes}
\label{sec:ag}

We now present \emph{algebraic-geometry codes} (``AG codes''), which are the main object used in this chapter,
and we show how they can give rise to diffusion matrices with interesting parameters.
We introduce a few geometry notions along the way in order to be able to faithfully describe the codes we will
be using. 
We refer the reader to the relevant chapters of \cite{vanlint,tvn,stichtenoth,fulton} for a much more detailed and rigorous treatment of the subject. 

\paragraph{AG codes as a generalization of RS codes.}
We first give an intuition of the construction used in AG codes. We have seen that RS codes are MDS
and that they can be instantiated for many lengths and dimensions. One limitation of the construction
is that the maximal length of the code is limited by the cardinality of the alphabet $\Fq$: we
cannot evaluate the functions on more points than there are elements in $\Fq$ without introducing some
repetition; we can still slightly increase this maximal length by one (two for some
values of $k$ and $q$) by defining an \emph{extended} RS code, which is still MDS. However, by the MDS
conjecture, we do not expect to be able to do better than that.
Yet we may want to
have a tradeoff between the maximal length of a code construction and its \emph{designed minimal distance} $d^* \leq d$.

A natural way to increase the maximal length of an evaluation code is to consider functions over domains
of higher dimension than the ``line'' used in RS codes. For instance, one could take $\dom$ to be the
$m$-dimensional space $\Fq^m$, and $\funcspace$ to be the $m$-variate polynomials of $\Fq[x_0,\ldots,x_{n-1}]$
less than a certain degree $r$. This defines \emph{Reed-Muller codes} (``RM codes'') of order $r$.
Although this construction successfully yields codes longer than RS codes, the parameters of RM codes
are not always easy to find.

An alternative approach to construct longer variants of RS codes is the one followed by AG codes. Seeing
RS codes as working with a line of genus zero, the idea is to take as evaluation domains
the points of (plane, affine) curves of higher genera, which can be assimilated to a subset of $\Fq^2$. A geometrically
meaningful way to choose the function space $\funcspace$ is then to consider certain subspaces of
the function fields of the curves. We will see that for well-chosen $\funcspace$, this construction
leads to codes of parameters $[n, k, d \geq n - k + 1 - g]_{\Fq}$ for curves of genus $g$. As the
(maximal) number of points on a curve is roughly increasing with the genus, this construction gives us the
sort of tradeoff we were searching for.

\subsection{Selective algebraic geometry notions}

We will be needing a few tools from algebraic geometry to make our description of AG codes explicit. We only introduce
here the notions that are relevant to this specific application,
largely ignoring the wider picture. Note that unlike the remainder of this chapter, the underlying fields used in the definitions
are not necessarily finite.

As we see AG codes as evaluation codes, we need to be able to: 1) Describe the evaluation domains.
2) Define the function spaces $\funcspace$ associated with given evaluation domains,
and be able to compute their bases.

\medskip

The construction of the examples of AG codes that we later consider can eventually be described in a purely affine way. However, in order to
justify them, we need to be able to reason about points at infinity. Thus we start with the following:

\begin{defi}[Affine space, projective space]
\label{def:affiprojec}
The $n$-dimensional \emph{affine space} $\affi^n(k)$ over a field $k$ is a set of \emph{points} formed by $n$-tuples over $k$,
$P \defas (x_0, \ldots, x_{n-1})$, $x_i \in k$.
The $n$-dimensional \emph{projective space} $\projec^n(k)$ over $k$ is a set of points formed by equivalence classes of $n+1$-tuples
$P \defas (x_0 : \ldots : x_n)$, $x_i \in k$, where not all $x_i$s are zero, and the equivalence relation $\sim$ is
defined by $(x_0 : \ldots : x_n) \sim (\lambda x_0 : \ldots : \lambda x_n)$, $\lambda \in k^*$.

\noindent
$\affi^n(k)$ can be embedded into $\projec^n(k)$ in a natural way as $(x_0, \ldots, x_{n-1}) \mapsto (x_0 : \ldots : x_{n-1} : 1)$.
The set of projective points $(x_0 : \ldots : x_{n-1} : 0)$ is called the \emph{hyperplane at infinity}.
\end{defi}

We now define projective varieties.

\begin{defi}[Projective variety]
A \emph{projective variety} in the projective space $\projec^n(k)$ over an algebraically closed field $k$ is the set of common zeros
of a prime ideal $\mathcal{I} \defas \{f_0, \ldots, f_m\}$ of homogeneous polynomials of $k[x_0,\ldots,x_n]$ (a polynomial is called \emph{homogeneous}
if all its monomials have the same degree).
\end{defi}

A \emph{projective curve} is a projective variety of dimension one. In practice, we will only consider curves in $\projec^2$ (plane curves),
which will then be defined by a single homogeneous polynomial $E$.

\paragraph{Remarks.}
\begin{enumerate}
\item The condition that $k$ be algebraically closed is not as restrictive as it may seem. Say that we wish to work in $\Fq$ (meaning in particular
that all coefficients of $E$ are in $\Fq$); though we will define a curve $\curve$ over the algebraic closure of $\Fq$, we
will only be interested in $\curve(\Fq)$, the points of $\curve$ that lie in $\Fq$ (its \emph{$\Fq$-rational points}, or simply rational points). 
\item For a given projective variety, say $\curve$ defined by $E(x,y,z)$, one can consider the corresponding affine variety defined as the zeros of ``$E(x,y,1)$,
$E \in k[x,y]$'' (the ``dehomogenised'' of $E$). It will have the same points (in the sense of the embedding of \autoref{def:affiprojec}) as $\curve$, minus its potential points at infinity.
Similarly, an affine variety can be extended to its projective closure.
\end{enumerate}

Our first step towards defining suitable function spaces $\funcspace$ is to define the function field of a variety.
We give two definitions, one in the affine and one in the projective case.

\begin{defi}[Coordinate ring, function field of an affine variety]
Let $\curve$ be an affine variety in $\affi^n(k)$ defined by the ideal $\mathcal{I}$. The \emph{coordinate ring} $k[\curve]$ of $\curve$ is $k[x_0, \ldots, x_{n-1}]/\mathcal{I}$.
The \emph{function field} $k(\curve)$ of $\curve$ is the quotient field of $k[\curve]$.
\end{defi}

\begin{defi}[Function field of a projective variety]
Let $\curve$ be an affine variety in $\projec^n(k)$ defined by the ideal $\mathcal{I}$. Let $F$, $G$ be homogeneous polynomials of equal degrees in $k[x_0, \ldots, x_n]$
with $G \notin \mathcal{I}$. The \emph{function field} $k(\curve)$ of $\curve$ is the set of \emph{rational functions} $F/G$ modulo $\mathcal{I'} \defas \{F/G~|~ F \in \mathcal{I}\}$.

\noindent
A rational function $F/G$ is \emph{regular} at $P$, point of $\curve$, if $G(P) \neq 0$.
\end{defi}

We will need to be able to talk about the behaviour of functions at specific points of a curve. For this, we first use the following:

\begin{defi}[Local ring of a point]
The \emph{local ring} $\locring_P$ of a point $P$ of a variety $\curve$ is the set of rational functions regular at $P$.
It has a unique maximal ideal $\maxi_P \defas \{f \in \locring_P ~|~ f(P) = 0\}$. 
\end{defi}

The local ring and its maximal ideal are in particular useful to define the \emph{tangent space} at a point $P$, which in turn allows to define smooth and singular points,
depending on its dimension.
We will not give the general definition here, mentioning instead  the following convenient alternative characterization in the specific case of curves:

\begin{defi}[Smooth and singular points of a curve, local parameter]
A point $P$ of a curve $\curve$ is \emph{smooth} or \emph{non-singular} iff $\maxi_P$ is a principal ideal. Any generator of the ideal ($t_P \in \locring_P$ s.t. $\maxi_P = t_P\locring_P$)
is called a \emph{local parameter} at $P$.

\noindent
A point that is not smooth is called \emph{singular}.
\end{defi}

To build an AG code from a curve $\curve$, we will require that it is smooth, \ie that none of its points are singular.
% (actually, we will
%only require this for $\curve(\Fq)$ for some $\Fq$).
We only consider smooth curves until \autoref{sec:nonsingmod}.

The local parameters are useful in defining the order of zeros and poles of arbitrary rational functions at a given point.
Let $f \neq 0$ be a rational function regular at $P$ and $t_P$ be a local parameter at $P$. The \emph{order} of $f$ at $P$ is:
\[
\ord_P(f) \defas \max \{k~ |~ f \in t_P^k\locring_P,~ f \notin t_P^{k+1}\locring_P\}.
\]
This can be extended to arbitrary rational functions by writing them as quotients of functions of $\locring_P$:
\[
\ord_P(f/g) = \ord_P(f) - \ord_P(g),~f,g\in\locring_P.
\]
For a function $f$, if $\ord_P(f) \geq 0$, we call this value the \emph{zero order} of $f$ at $P$; if $\ord_P(f) < 0$
we call $-\ord_P(f)$ the \emph{pole order} of $f$ at $P$.

We note in particular the two useful properties of $\ord$:
1) $\ord(fg) = \ord(f) + \ord(g)$; 2) $\ord(1/f) = -\ord(f)$.

%We will shortly see that being able to compute the order of a function at some desired points is essential in the definition of $\funcspace$.
In effect, $\ord_P$ defines a \emph{discrete valuation} of functions of $k(\curve)$, associated with the point $P$, \ie $\locring_P$ is a \emph{valuation ring}:

\begin{defi}[Valuation ring]
A ring $\locring$ is a \emph{valuation ring} of $k(\curve)$ if $k \subset \locring \subset k(\curve)$  and for every $f \in k(\curve)$,
$f \in \locring$ or $f^{-1} \in \locring$.
\end{defi}

We will see shortly that the ability to define valuations at any point is essential
in the definition of $\funcspace$; this is why we require the curve to be non-singular, as a local ring is a valuation ring iff $P$ is smooth.

\medskip

The last main objects that we introduce are the divisors on a curve.

\begin{defi}[Divisor]
A \emph{divisor} on a smooth curve $\curve$ is a formal (finite) sum of points $P_i$ of $\curve$: $D \defas \sum a_iP_i$, $a_i \in \mathbf{Z},~P_i \in \curve$.
The \emph{support} of a divisor is the set of points $P_i$ with $a_i \neq 0$.

\noindent
The set $\cdiv(\curve)$ of divisors on $\curve$ with formal addition and negation of divisors ($\sum a_iP_i + \sum b_iP_i \defas \sum (a_i+b_i)P_i$;
$-\sum a_iP_i \defas \sum -a_iP_i$) forms a group.

\noindent
A divisor $D \defas \sum a_iP_i$ is called \emph{effective} ($D \geq 0$) if $\forall i,~a_i \geq 0$.

\noindent
The \emph{degree} $\deg(D)$ of $D = \sum a_iP_i$ is defined as $\sum a_i$
\end{defi}

\begin{defi}[Divisor of a rational function (principal divisor)]
The divisor $(f)$ of a rational function $f \in k(\curve)$, $f \neq 0$, is defined as $\sum \ord_P(f)P$.
A divisor $D$ equal to $(f)$ for some function $f$ is called a \emph{principal divisor}.
\end{defi}

Principal divisors form a subgroup of $\cdiv(\curve)$. It can be shown that they are always of degree zero
(this is equivalent to saying that a rational function always has the same number of poles and zeros, counted with
multiplicity).

\begin{defi}[Space associated with a divisor]
Let $D \in \cdiv(\curve)$. We define $\rrs(D)$, the \emph{space associated with $D$} (or \emph{Riemann-Roch space}) as:
\[
\rrs(D) = \{f \in k(\curve)^*~|~(f) + D \geq 0\} \cup \{0\}.
\]
This forms a vector space; we write $\ell(D)$ its dimension.
\end{defi}

This is finally the kind of function space we will be using to define AG codes. For a divisor
$D \defas \sum a_iP_i - \sum b_jP_j$, $a_i, b_j \geq 0$, a simple reformulation of $\rrs(D)$ is to say that it consists of the zero function and of all the functions
which may have poles only in the $P_i$s of order at most $a_i$, and which must have zeros in all the $P_j$s of order
at least $b_j$.

It can be shown that for any $D$, $\rrs(D)$ is finite-dimensional. In fact, we have the following major theorem from Riemann and Roch, which allows to compute the explicit value $\ell(D)$
for a given divisor $D$.

\begin{thm}[Riemann-Roch theorem]
Let $D$ be a divisor on a curve of \emph{genus} $g$, then $\ell(D) \geq \deg(D) + 1 - g$, with equality
when $\deg(D) > 2g - 2$.
\end{thm}

The formulation of this theorem uses the notion of \emph{genus} of a curve, that we have not defined yet. A possible, purely algebraic definition is precisely to take
$g(\curve) \defas \max \{\deg(D) - \ell(D) + 1,~D \in \cdiv(\curve)\}$. Although this is not very satisfying, we will not provide a better definition here (and computing
the genus in this way is not as hard as it may seem); in practice, the genus
of the curves we use will be known beforehand.

\subsubsection{Working with singular curves}
\label{sec:nonsingmod}

In the last few paragraphs, we have assumed that we were working with a smooth curve. We justified this by the need of being able to define a valuation for functions
at any point of the curve, which allowed us to define divisors and their associated vector spaces.
Unfortunately, some of the curves we may wish to use in practice are not smooth. We briefly and informally sketch here how it is possible to nonetheless work with some
of these curves. We refer the reader to \eg Fulton~\cite[Chap. 7]{fulton} for a serious treatment of the matter.

A first remark is that what effectively matters in the definition of the function space is not so much the points on the curves as their associated valuation
rings. It is in fact possible to define entirely what we have presented above in a purely algebraic way, as is done for instance by Stichtenoth~\cite{stichtenoth}.
In this case, the notion of points is replaced by the one of \emph{places} of a function field (written $F/k$, not necessarily associated with a curve), where a place is simply the (necessarily unique) maximal
ideal of a valuation ring of the function field:

\begin{defi}[Place of a function field]
A place $P$ of a function field $F/k$ is the maximal ideal of a a valuation ring $\locring$ of $F/k$. 
\end{defi}

Divisors can then be redefined as sums of places instead of sums of points.

For a smooth curve $\curve$, there is a bijection between its points and the valuation rings (the places) of $k(\curve)$, and the theory as presented above is sufficient.
On the other hand, if $\curve$ is singular, we would like to be able to find a way of defining places at (some of) its singularities, on top of the smooth points.
This can be done by constructing a \emph{non-singular model} $\nscurve$ of $\curve$,
and taking as places of $\curve$ the points of $\nscurve$. A useful theorem is that for a projective curve $\curve$, there is a unique (up to isomorphism) non-singular curve
$\nscurve$ that is birationally equivalent to it\footnote{This means that there are \emph{rational maps} from $\curve$ to $\nscurve$ and vice-versa, that compose to the identity wherever they are
defined.
A rational map $\curve \subseteq \projec^m \rightarrow \projec^n$ is an $n+1$-tuple of homogeneous polynomials of degree $m+1$ s.t. the polynomials do not jointly
vanish at any point of the curve.}. Furthermore, the function fields $k(\curve)$ and $k(\nscurve)$ are the same, up to isomorphism, so defining the places of $\curve$ in this
way is meaningful.
(We will not address here the problem of computing these non-singular models \emph{per se}.)

The map $\curve \rightarrow \nscurve$ may not be defined on every point of $\curve$, but the number of points where it is not defined is finite (this makes sense, as a curve
has at most finitely many singular points). We say that a place is \emph{centered} at a point $P$ if $P$ is its image by the map $\nscurve \rightarrow \curve$. Any non-singular
point of $\curve$ has exactly one place centered at it, while finitely many places may be centered at singular points.

We conclude by defining curves that are said to be in \emph{special position} (cf. \eg \cite{DBLP:journals/tit/SaintsH95}), which are convenient to use for several reasons.

\begin{defi}[Curve in special position]
A projective plane curve $\curve$ is in \emph{special position} if:
\begin{enumerate}
\item it has exactly one point $P_\infty$ at infinity;
\item there is exactly one place centered at $P_\infty$;
\item the affine curve $\curve_A \defas \curve\backslash P_\infty$ is smooth;
\item the pole order of $x/z$ and $y/z$ at $P_\infty$ are not equal.
\end{enumerate}
\end{defi}

Curves in special positions are useful as there is a bijection between their places and their points (even if they are potentially singular); thus they can always be considered
``as if smooth'' w.r.t. to the discussions of this section. Additionally, if one considers the space $\rrs(aP_\infty)$, $a > 0$ of functions having poles only at the point at infinity,
its elements can be mapped to polynomials of the coordinate ring of $\curve_A$. In other words, one can perform all the computations in $\rrs(aP_\infty)$ with the affine curve
only (we discuss how to explicitly compute bases for such spaces next). The main curve used in this chapter is a singular curve in special position.

\subsubsection{Computation of a basis of $\rrs(aP_\infty)$ for a curve in special position}

We conclude this brief presentation of algebraic curves and their function fields by showing how to compute a basis of $\rrs(aP_\infty$), $a > 0$
for a plane curve in special position.

Consider $\curve$ defined by a homogeneous equation $E(x,y,z)$. From a local parameterization at the place $P_\infty$, we can compute the pole order of
$f \defas x/z$ and $g \defas y/z$ at $P_\infty$ (which is enough to determine the pole
order of any fraction of $x$, $y$ and $z$).
An element of $\rrs(aP_\infty)$ is a linear combination $\sum_\alpha \alpha f^{\alpha_i}g^{\alpha_j}(z/z)^{\alpha_k}$ s.t. $\alpha_i + \alpha_j + \alpha_k$ is constant (the function is
homogeneous) and $\forall \alpha,~\alpha_i\ord_{P_\infty}(f) + \alpha_j\ord_{P_\infty}(g) \leq a$ (the pole order at $P_\infty$ of the monomials is less
than $a$).
A generating family for $\rrs(aP_\infty)$ is thus (omitting the homogenising terms $(z/z)^{\alpha_k}$ for convenience) $\{f^ig^j~|~i\times\ord_{P_\infty}(f) + j\times\ord_{P_\infty}(g) \leq a\}$; if we keep only a single monomial
with any given pole order, it forms a basis.
As $P_\infty$ is the only place at infinity, we can switch to the affine case for all computations by considering points in affine coordinates and dehomogenising $E$ and the basis functions; notably, the latter effectively become polynomials rather
than rational fractions. Indeed, a basis for $\rrs(aP_\infty) \subseteq k[\curve]$ is formed by $\{x^iy^j~|~i\times\ord_{P_\infty}(f) + j\times\ord_{P_\infty}(g) \leq a\}$.

\medskip

We conclude by presenting a simple technique to compute the pole order of $x/z$ and $y/z$ at the place at infinity for a curve in special position.
In this case, we know by definition that the valuation $\ord_{P_\infty}$ is well defined and that it is negative and distinct for $x/z$ and $y/z$.
Thus $\exists \alpha,\beta \in \mathbf{N}^*$ s.t. $(x/z)^\beta(y/z)^{-\alpha}$ is regular and non-zero at $P_\infty$. The smallest such $\alpha$
and $\beta$ are the pole order of $x/z$ and $y/z$ respectively.

We give two examples of a computation with this technique, corresponding to the two curves used in the next \autoref{sec:constr_ag}.

\begin{example}[$x^2z + xz^2 = y^3+yz^2$]
\label{ex:order_gen1}
We consider the curve of equation $E(x,y,z) := x^2z + xz^2 = y^3+yz^2$ over a field of characteristic two. It has a unique place at infinity $P_\infty \defas [1:0:0]$.
From the equation of the curve, we may guess 3 and 2 to be the orders of $x/z$ and $y/z$, as the fraction $x^2z^3/y^3z^2$ seems to be the one of least degree
that can be simplified using $E$.

We may then rewrite $x^2z^3$ in the four following ways:
\begin{enumerate}
\item $z^2(x^2z)$;
\item $z^2(y^3 + yz^2 + xz^2)$;
\item $xz(xz^2)$;
\item $xz(y^3 + yz^2 + x^2z)$.
\end{enumerate}
Similarly, $y^3z^2$ can be rewritten as:
\renewcommand{\theenumi}{\roman{enumi}}
\begin{enumerate}
\item $z^2(y^3)$;
\item $z^2(x^2z + xz^2 + yz^2)$;
\item $y^2(yz^2)$;
\item $y^2(y^3 + x^2z + xz^2$.
\end{enumerate}
We can then notice that ``1/ii'' = $z^2(x^2z)/z^2(x^2z + xz^2 + yz^2)$ simplifies to $x^2z/(x^2z + xz^2 + yz^2)$ = $x^2/(x^2 + xz + yz)$,
which evaluates to 1 at $[1:0:0]$. Hence we indeed have $\ord_{P_\infty}(x/z) = 3$ and $\ord_{P_\infty}(y/z) = 2$.
\end{example}

\begin{example}[$x^5 = y^2z^3 + yz^4$]
\label{ex:order_gen2}
We consider the curve of equation $E(x,y,z) := x^5 = y^2z^3 + yz^4$ over a field of characteristic two. It has a unique place at infinity $P_\infty \defas [0:1:0]$.
As in \autoref{ex:order_gen1}, we may guess that $x/z$ and $y/z$ have order 2 and 5 respectively. We thus consider the fraction $x^5z^2/y^2z^5$.
This can be rewritten as $(y^2z^5 + yz^6)/y^2z^5$ which readily simplifies to $(y^2 +yz)/y^2$ = $(y+z)/y$ which evaluates to 1 at $[0:1:0]$.
Hence we indeed have $\ord_{P_\infty}(x/z) = 2$ and $\ord_{P_\infty}(y/z) = 5$.
\end{example}

\subsection{Construction of AG codes}
\label{sec:constr_ag}

We are now ready to define the class of AG codes that we will use in the remainder of this work.

Let $\curve$ be a smooth projective curve defined by an equation in $\Fq$ such that the set of its rational places $\curve(\Fq)$ is non-empty.
Let $D \in \cdiv(\curve)$ be a divisor on $\curve$ and $\points \defas \langle P_0, \ldots, P_{n-1}\rangle \subset \curve(\Fq)$ be $n$ places not in the support of $D$.
This defines a code $\evcode(\rrs(D), \points)$.

For the definition to be valid, the evaluation map from $\rrs(D)$ to $\points$ has to be injective (\ie have only $\{0\}$ as kernel).
This condition can easily be expressed in terms of divisors; if we call $S \defas P_0 + \ldots + P_{n-1}$ the divisor of the
evaluation support, we need $\rrs(D - S)$ (\ie functions of $\rrs(D)$ with zeros in all of the $P_i$s, \ie the kernel of $\evmap$) to be only the zero function.
One can see that this will always be the case as long as $\deg(D) < \deg(S)$.

\medskip

Assuming that we defined a code using this construction, one of our main concerns would be to determine its parameters $k$ and $d$ (or at least provide an estimate).
By definition of the minimal distance, $n - d$ is the maximal number of zeros that a function of $\rrs(D)$ can have
over $\points$; we already showed that the latter is upper-bounded by $\deg(D)$, so we have $d \geq n - \deg(D)$. This lower-bound
$n - \deg(D)$ for $d$ is the designed minimum distance $d^*$ for the code.

The dimension of the code is simply $\ell(D)$, which can be computed from the Riemann-Roch theorem. In particular, for $\deg(D) > 2g -2$, $g$ the genus of
$\curve$, we have $k = \ell(D) = \deg(D) + 1 - g$. We can then rephrase the lower-bound on $d$ in terms of $k$ as $d \geq n - k + 1 - g$, which is indeed
what we promised at the beginning of the section.

\subsubsection{Examples}

We give two concrete examples of AG codes, the second of which being the code used to define the diffusion matrices presented in this chapter.

\begin{example}[An AG code from a curve of genus 1]
\label{ex:genus1}
Let $\curve$ be the plane projective curve of genus 1 of equation $x^2z + xz^2 = y^3 + yz^2$. It is smooth, in special position,
with a unique point at infinity $P_\infty \defas [1:0:0]$, and has 24 $\Fst$-rational affine points. The pole order
of $x/z$ and $y/z$ at $P_\infty$ is 3 and 2 respectively (as computed in \autoref{ex:order_gen1}).

Let $D \defas 12P_\infty$. From the Riemann-Roch theorem, the dimension $\ell(D)$ of $\rrs(D)$ is $12 + 1 - g = 12$. We can thus
define a $[24, 12, 12 \geq]_{\Fst}$ AG code $\evcode(\rrs(D), \points)$ with $\points$ the affine rational points of $\curve$
put in any order.

This code can be used to define a diffusion matrix of dimension 12 over $\Fst$, which diffuses over $12\times 4 = 48$ bits and
has differential (and in fact also linear) branch number at least 12 (in fact exactly 12). 
\end{example}

\begin{example}[An AG code from a curve of genus 2]
\label{ex:genus2}
Let $\curve$ be the plane projective curve of genus 2 of equation $x^5 = y^2z^3 + yz^4$. It is singular, in special position,
with a unique point at infinity $P_\infty \defas [0:1:0]$ at which a unique place is centered. It has 32 $\Fst$-rational affine points. The pole order
of $x/z$ and $y/z$ at $P_\infty$ is 2 and 5 respectively (as computed in \autoref{ex:order_gen2}).

Let $D \defas 17P_\infty$. From the Riemann-Roch theorem, the dimension $\ell(D)$ of $\rrs(D)$ is $17 + 1 - g = 16$. We can thus
define a $[32, 16, 15 \geq]_{\Fst}$ AG code $\evcode(\rrs(D), \points)$ with $\points$ the affine rational points of $\curve$
put in any order.

This code can be used to define a diffusion matrix of dimension 16 over $\Fst$, which diffuses over $16\times 4 = 64$ bits and
has differential (and in fact also linear) branch number at least 15 (in fact exactly 15). This code is also self-dual,
which implies in particular that its derived diffusion matrices are orthogonal.
\end{example}
